R Packages to Aid in Handling Web Access Logs

### 2015-12-03 paper/paper.Rmd
add benchmarks

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

@@

This section may contain a figure such as Figure \ref{figure:rlogo}.

@@

@@

There will likely be several sections, perhaps including code snippets, such as:

```{r}
x <- 1:10
x
```

>>>>

The rise of the World Wide Web over the last few decades has made it dramatically easier to access and transfer data, and R \citep{RCore} boasts abundant functionality when it comes to taking data *from* the web. Base R itself has simple file downloading and page reading capabilities, through the `download.file` and `readLines` functions, and additional functionality is made available for handling web-accessible data through packages such as **httr** \citep{httr}.

Data *on* the web is not, however, the only kind of web data that interests researchers; web traffic is, in and of itself, an interesting data source. Access logs - records of connections between users and a web server - are an asset and resource for people studying everything from user behaviour on the internet \citep{halfak}, to website performance \citep{performance}, to information security \citep{infosec}.

As a statistically-oriented programming language, R is commonly used by these same researchers for data analysis, testing and reporting - but it lacks tools designed for the kinds of data sources and formats that are encountered when dealing with access logs. In this article we review the use cases for particular operations over web data, the limitations in base R when it comes to performing those operations, and a suite of R packages designed to overcome them: **webreadr** \citep{webreadr}, designed for reading access logs in,
**urltools** \citep{urltools}, for URL manipulation, **iptools** \citep{rgeolocate} for handling IP addresses, and **rgeolocate** \citep{rgeolocate} for direct IP geolocation.

@@

The first task with any data analysis is to read the data into R so that it can be manipulated. With access logs this is slightly complicated by the fact that there is no one standard for what a log should look like; instead, there are multiple competing approaches from different software platforms and eras. These include the Common Log Format (CLF), the confusingly-named Combined Log Format, and formats used by individual, commonly-used software platforms - such as the custom format for the Squid internet caching software, and the format used by Amazon Web Services (AWS).

The difference between formats can easily be shown by looking at how timestamps are represented:

\begin{table}[ht]
\centering
\caption{Timestamps in Common Access Log Formats}
\begin{tabular}{rlr}
  \hline
Log Type & Timestamp Columns & Timestamp Format \\ 
  \hline
Common Log Format & 1 & 10/Oct/2000:13:55:36 -0700\\ 
Combined Log Format & 1 & 26/Apr/2000:00:23:48 -0400\\ 
Squid & 1 & 1286536309.450\\ 
AWS & 2 & 2014-05-23	01:13:11\\ 
   \hline
\end{tabular}
\end{table}

With four log types, we have three different timestamp formats - and that's only one of the many columns that could appear. These logs also vary in whether they specify quoting fields (or sanitising unquoted ones), the columns they contain and the data each column contains in turn.

To make reading access logs into R as easy as possible we created the **webreadr** package. This contains user-friendly equivalents to `read.table` for each type of log, detecting the fields that should appear, converting the timestamps into POSIX objects, and merging fields or splitting fields where necessary. The package contains four core functions, one for each form of log: `read_clf` for the Common Log Format, `read_combined` for the Combined Log Format, and `read_squid` and `read_aws` for Squid and AWS formats respectively. Each one abstracts away the complexity of specifying column names and formats, and instead allows a researcher to read a file in with a minimal amount of work.

As the name suggests, it is built not on top of base R but on top of the **readr** \citep{readr} package, allowing us to take advantage of substantial speed improvements that package's base functions have over base R. These improvements can be seen in the visualisation below, which uses **microbenchmark** \citep{microbenchmark} to compare 100 reads of a 600,000-line "squid" formatted file with webreadr to the same operation performed in base R:

@@

**webreadr**

@@

Within access logs, URLs appear - both to describe the web asset or page that the user requested and where the user came from. These fields are usually named <code>url</code> and <code>referer</code> respectively.

### Decoding
Both values can be percent-encoded, allowing them toinclude characters that are valid but reserved by the URL specification as having special meanings ("reserved characters"). A `#` symbol, for example, is encoded as <code>%23</code>: a percentage symbol, followed by the byte-code for that character.

The encoding of reserved characters is useful, since it means that URL paths and queries can contain a vast array of values - but it makes data analysis tougher to do. Examples of common data analysis or cleaning operations that become more difficult are:

1. **Aggregation**. Aggregating URLs together is useful to identify, for example, the relative usage and popularity of particular pages
in your data - but it becomes tougher if encoding is inconsistent, because two URLs could hold the same value but *look* very different.
2. **Value selection**. With text-based data, regular expressions are a common way of filtering or selecting entries that meet particular conditions, but things become fuzzy when you have to look not just for particular characters (a space, say) but also the encoded values (<code>%20</code>).
3. **Exploratory data analysis** (EDA). EDA is a common initial step to investigate a data set, examining the variables and values it contains and whether they meet a researcher's expectations - but on a practical basis it becomes difficult when the values aren't human-readable.

The solution is to be able to consistently decode URLs, which makes URL-based data far easier to analyse. Base R does contain a function, <code>URLdecode</code>, for doing just that, but is neither vectorised nor based on compiled code; with large datasets it takes an extremely long time.

To solve this common problem in analysing request logs, the **urltools** [@urltools] package was created. This contains a function,
<code>url_decode</code>, which decodes URLs - but relies on vectorised, compiled code. Benchmarking the two approaches against each other shows that the `urltools` implementation is approximately 60-70 times faster over large datasets. Again using **microbenchmark**, if we compare the vectorised decoding of 1,000,000 URLs with `urltools` against `URLdecode` in a `vapply` loop, we see a 60-70 times speed improvement of urltools over base R:

### 2015-12-03 paper/paper.tex
add benchmarks

\abstract{
Web access logs contain information on HTTP(S) requests and form a key
part of both industry and academic explorations of human behaviour on
the internet - explorations commonly performed in R. In this paper we
explain and demonstrate a series of packages - \emph{webreadr},
\emph{urltools}, \emph{iptools} and \emph{rgeolocate} - designed to
efficiently read in, parse and munge access log data, allowing
researchers to handle it easily.
}

@@

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

\subsection{Section title in sentence
case}\label{section-title-in-sentence-case}

This section may contain a figure such as Figure \ref{figure:rlogo}.

@@

\subsection{Another section}\label{another-section}

There will likely be several sections, perhaps including code snippets,
such as:

\begin{Schunk}
\begin{Sinput}
x <- 1:10
x
\end{Sinput}
\begin{Soutput}
#>  [1]  1  2  3  4  5  6  7  8  9 10
\end{Soutput}
\end{Schunk}

>>>>

\abstract{
Web access logs contain information on HTTP(S) requests and form a key
part of both industry and academic explorations of human behaviour on
the internet - explorations commonly performed in R. In this paper we
explain and demonstrate a series of packages - \textbf{webreadr},
\textbf{urltools}, \textbf{iptools} and \textbf{rgeolocate} - designed
to efficiently read in, parse and munge access log data, allowing
researchers to handle it easily.
}

@@

The rise of the World Wide Web over the last few decades has made it
dramatically easier to access and transfer data, and R \citep{RCore}
boasts abundant functionality when it comes to taking data \emph{from}
the web. Base R itself has simple file downloading and page reading
capabilities, through the \texttt{download.file} and \texttt{readLines}
functions, and additional functionality is made available for handling
web-accessible data through packages such as \textbf{httr} \citep{httr}.

Data \emph{on} the web is not, however, the only kind of web data that
interests researchers; web traffic is, in and of itself, an interesting
data source. Access logs - records of connections between users and a
web server - are an asset and resource for people studying everything
from user behaviour on the internet \citep{halfak}, to website
performance \citep{performance}, to information security
\citep{infosec}.

As a statistically-oriented programming language, R is commonly used by
these same researchers for data analysis, testing and reporting - but it
lacks tools designed for the kinds of data sources and formats that are
encountered when dealing with access logs. In this article we review the
use cases for particular operations over web data, the limitations in
base R when it comes to performing those operations, and a suite of R
packages designed to overcome them: \textbf{webreadr} \citep{webreadr},
designed for reading access logs in, \textbf{urltools} \citep{urltools},
for URL manipulation, \textbf{iptools} \citep{rgeolocate} for handling
IP addresses, and \textbf{rgeolocate} \citep{rgeolocate} for direct IP
geolocation.

\subsection{Reading access logs}\label{reading-access-logs}

The first task with any data analysis is to read the data into R so that
it can be manipulated. With access logs this is slightly complicated by
the fact that there is no one standard for what a log should look like;
instead, there are multiple competing approaches from different software
platforms and eras. These include the Common Log Format (CLF), the
confusingly-named Combined Log Format, and formats used by individual,
commonly-used software platforms - such as the custom format for the
Squid internet caching software, and the format used by Amazon Web
Services (AWS).

The difference between formats can easily be shown by looking at how
timestamps are represented:

\begin{table}[ht]
\centering
\caption{Timestamps in Common Access Log Formats}
\begin{tabular}{rlr}
  \hline
Log Type & Timestamp Columns & Timestamp Format \\ 
  \hline
Common Log Format & 1 & 10/Oct/2000:13:55:36 -0700\\ 
Combined Log Format & 1 & 26/Apr/2000:00:23:48 -0400\\ 
Squid & 1 & 1286536309.450\\ 
AWS & 2 & 2014-05-23    01:13:11\\ 
   \hline
\end{tabular}
\end{table}

With four log types, we have three different timestamp formats - and
that's only one of the many columns that could appear. These logs also
vary in whether they specify quoting fields (or sanitising unquoted
ones), the columns they contain and the data each column contains in
turn.

To make reading access logs into R as easy as possible we created the
\textbf{webreadr} package. This contains user-friendly equivalents to
\texttt{read.table} for each type of log, detecting the fields that
should appear, converting the timestamps into POSIX objects, and merging
fields or splitting fields where necessary. The package contains four
core functions, one for each form of log: \texttt{read\_clf} for the
Common Log Format, \texttt{read\_combined} for the Combined Log Format,
and \texttt{read\_squid} and \texttt{read\_aws} for Squid and AWS
formats respectively. Each one abstracts away the complexity of
specifying column names and formats, and instead allows a researcher to
read a file in with a minimal amount of work.

As the name suggests, it is built not on top of base R but on top of the
\textbf{readr} \citep{readr} package, allowing us to take advantage of
substantial speed improvements that package's base functions have over
base R. These improvements can be seen in the visualisation below, which
uses \textbf{microbenchmark} \citep{microbenchmark} to compare 100 reads
of a 600,000-line ``squid'' formatted file with webreadr to the same
operation performed in base R:

@@

\textbf{webreadr}

\subsection{Decoding and parsing URLs}\label{decoding-and-parsing-urls}

Within access logs, URLs appear - both to describe the web asset or page
that the user requested and where the user came from. These fields are
usually named url and referer respectively.

\subsubsection{Decoding}\label{decoding}

Both values can be percent-encoded, allowing them toinclude characters
that are valid but reserved by the URL specification as having special
meanings (``reserved characters''). A \texttt{\#} symbol, for example,
is encoded as \%23: a percentage symbol, followed by the byte-code for
that character.

The encoding of reserved characters is useful, since it means that URL
paths and queries can contain a vast array of values - but it makes data
analysis tougher to do. Examples of common data analysis or cleaning
operations that become more difficult are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \textbf{Aggregation}. Aggregating URLs together is useful to identify,
  for example, the relative usage and popularity of particular pages in
  your data - but it becomes tougher if encoding is inconsistent,
  because two URLs could hold the same value but \emph{look} very
  different.
\item
  \textbf{Value selection}. With text-based data, regular expressions
  are a common way of filtering or selecting entries that meet
  particular conditions, but things become fuzzy when you have to look
  not just for particular characters (a space, say) but also the encoded
  values (\%20).
\item
  \textbf{Exploratory data analysis} (EDA). EDA is a common initial step
  to investigate a data set, examining the variables and values it
  contains and whether they meet a researcher's expectations - but on a
  practical basis it becomes difficult when the values aren't
  human-readable.
\end{enumerate}

The solution is to be able to consistently decode URLs, which makes
URL-based data far easier to analyse. Base R does contain a function,
URLdecode, for doing just that, but is neither vectorised nor based on
compiled code; with large datasets it takes an incredibly long time.
